NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-06-22 Manuscript/manuscriptPS.tex
Update texrt

@@

\subsection{Headgroup structure in PS and PC mixtures}
Dilution of PS lipid bilayers with PC lipids reduces propensity for the formation of
strong complexes with multivalent ions and is proposed to make PS headgroups 
less rigid~\cite{browning80,buldt81,roux90,roux91}. Therefore, the intermolecular
interactions at the headgroup region seems to determine important physical properties 
of mixed lipid bilayers. The headgroup order parameters of POPC increase in experiments 
with increasing amount of POPS in the mixed lipid bilayer (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
This behaviour is generally observed when negatively charged lipids or surfactants 
are mixed with PC lipids~\cite{scherer87,scherer89} and can be understood by the 
tilting of lipid headgroup more parallel to the membrane plane according to
the electrometer concept \cite{seelig87}. The headgroup order parameters of PS lipids
shift closer to zero when bilayer is diluted with PC lipids in 
experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.

The increase of POPC headgroup order parameters with the increasing
amount of negatively charged POPS lipid is reproduced in
MacRog simulations with potassium counterions, but not in 
Berger simulations with sodium or in CHARMM36 simulations
with potassium or sodium conterions (Fig. \ref{HGorderparametersPCvsPS}). 
The observations can be explained using the electrometer concept.
The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
which surpasses the effect of negatively charged lipids as the amount of
counterions increase with increasing amount of PS.
In CHARMM36 simulations, the counterion binding neutralizes the effect
of PS and the headgroup order parameters are not changed with increasing amount
of PS. Finally, the weak binding of potassium in the MacRog simulations
enables the increase of order parameters with the added amount of PS 
(Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).

@@

%Also the Ca2+ binding affinity to bilayers containing
%PS lipids is measured from mixtures with PC, because pure PS
%bilayers exhibit instant transition to the ordered phase with the
%addition of the ions.
%Therefore, the characterization of mutual interactions
%between PC and PS lipid headgroups is necessary in order to
%understand PS lipids in biological environments.

The headgroup order parameter changes of POPC and POPS from POPC:POPS (5:1) mixtures
are shown in Fig. \ref{changesWITHCaClPS} as a function of Ca$^{2+}$ concentration from different
simulations and experiments \cite{roux90}. 
The ion density distributions from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
The results suggest that Ca$^{2+}$ ions clearly overbind in simulations with
MacRog and Berger models, as expected from previous results for PC lipid bilayers \cite{catte16}.
It should be noted, however, that the lowest concentration (100mM) gives
a good agreemet with experiments
\todo{Should be analyze/discuss this further?
  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
It should be noted that Berger simulation do not have counterions.}.
Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, incorporated
in the parameters from CHARMM-GUI at the time of running the simulations (January 2018).
The binding of calcium to pure POPC bilayer is also too weak with these parameters
as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
\todo{The discussion is to be finished when we have all the data in the plot.}

>>>>

\subsection{Comparison of ion binding between simulations and experiments using the electrometer concept}

@@

\subsection{Headgroup structure in PS and PC mixtures}
Dilution of PS lipid bilayers with PC lipids reduces the propensity
of PS headgroup-multivalent ion complexes and is proposed to make PS
headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
Therefore, the intermolecular interactions at the headgroup region seems
to be important for the physical properties of mixed lipid bilayers.
These interactions can be indirectly monitored by measuring the
headgroup order parameters from PS:PC mixtures with different molar
ratios. The headgroup order parameters of POPC increase in such experiments 
with increasing amount of POPS (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
This behaviour is generally observed when negatively charged lipids or surfactants 
are mixed with PC lipids~\cite{scherer87,scherer89} and can be understood by the 
tilting of lipid headgroup more parallel to the membrane plane according to
the electrometer concept \cite{seelig87}. The headgroup order parameters of PS lipids
shift closer to zero when bilayer is diluted with PC lipids in 
experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.

The increase of POPC headgroup order parameters with the increasing
amount of negatively charged POPS lipid is reproduced in
MacRog simulations with potassium counterions, but not in 
Berger simulations with sodium or in CHARMM36 simulations
with potassium or sodium conterions (Fig. \ref{HGorderparametersPCvsPS}). 
The observations can be explained using the electrometer concept.
The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
which surpasses the effect of negatively charged lipids as also the amount of
counterions increase with increasing amount of PS.
In CHARMM36 simulations, the counterion binding neutralizes the effect
of PS and the headgroup order parameters are not changed with increasing amount
of PS. Finally, the weak binding of potassium in the MacRog simulations
enables the increase of order parameters with the increasing amount of
negatively charged PS lipids (Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).

@@

Calcium ions form strong dehydrated molecular complexes with PS lipids,
which can also lead to phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
Such complexes and phase separation are not observed in
bilayers that are sufficiently diluted with PC
lipids \cite{feigenson86,mattai89,roux90,roux91}.
Therefore, the Ca2+ binding affinity to bilayers if often measured
from POPC:POPS (5:1) mixtures \cite{roux90,roux91}.

The headgroup order parameter changes of POPC and POPS from 
are shown in Fig. \ref{changesWITHCaClPS} as a function of Ca$^{2+}$ concentration from different
simulations and experiments \cite{roux90}. 
The ion density distributions from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
The results suggest that Ca$^{2+}$ ions clearly overbind in simulations with
MacRog and Berger models, as expected from previous results for PC lipid bilayers \cite{catte16}.
It should be noted, however, that the lowest concentration (100mM) gives
a good agreemet with experiments
\todo{Should be analyze/discuss this further?
  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
It should be noted that Berger simulation do not have counterions.}.
Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, incorporated
in the parameters from CHARMM-GUI at the time of running the simulations (January 2018).
The binding of calcium to pure POPC bilayer is also too weak with these parameters
as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
\todo{The discussion is to be finished when we have all the data in the plot.}
